Array

### 2014-02-18 DOCUMENTATION/AdvancedTypes.tex
All details about the translation of Array, included multidimensional and interval (shift support and cast types).

  \begin{tabular}[t]{rcl}
	\hline
        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
        \hline
        \B{op} & : & \B{BoolOp} \\
        \B{args} & : & seq \B{Pred} \\
        \hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{Struct} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Data structure : \B{Struct}} \\
        \hline
	\B{id} & : &  \B{Name}\\
        \B{elements} & : & seq \B{Name} \\
	\B{type} & : & seq \B{Type}\\
	\hline
%  	\multicolumn{3}{|c|}{Data record : \B{Record}} \\
%        \hline
%	\B{id} & : &  \B{Name}\\
%        \B{elements} & : & seq \B{Name} \\
%	\B{type} & : & \B{Struct}\\
%	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Expr}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Expr} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).
      Sequence and optional attributes are denoted ``seq'' and ``opt'',
      respectively}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

The element access of array can need a shift in indexes. When the domain of 
array is an interval and it is not started from zero, it must adjust the 
indexes shifting the access, because the addressing  of array in intermediate 
representation LLVM starts at 0.

% Rule to get a element in array, it can be used in become equal or Expressions
\begin{align*}
\lefteqn{\trad{n}{lvalue} \isdef} \\
& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}}"} \IN \\
& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
& \quad \IF \mbox{\B{n} is an array} \THEN  \trad{n}{ArrayItem} \\
& \quad \ELSE \LET v = \variable \IN \\
& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
\end{align*}

@@

A list of type and variable is generated by the following rule.
%TODO: insert a rule to do the shift address with the first element of interval
\begin{align*}
&\trad{n}{LRExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp}  \IN  \llvm{",\PH{t} \PH{v} "} \sep p\\
&\trad{n l}{LRExp} \isdef \LET r_1 \sep p_1 = \trad{n}{LRExp} \AND 
 \LET r_2 \sep p_2 = \trad{l}{LRExp} \IN\\
& \quad  \llvm{"\PH{r_1} \PH{r_2}"} \sep \llvm{"\PH{p_1} \PH{p_2}"}\\
\end{align*}

%TODO: support to shift between interval
\begin{itemize}
	\item $array2 \in (10..50) \times ((-10)..100) \to INT $   (a multidimensional array)
\end{itemize}

The array indexes can composed of only primitive types (INTEGER and BOOLEAN). To access an element the formula is $address(e(i1,..,in))  = i1-i1\_start, ... , in-in\_start$.

%TODO create a new rule based on the below rule.

%it must change the declaration and the rule of ArrayItem

>>>>

  \begin{tabular}[t]{rcl}
	\hline
        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
        \hline
        \B{op} & : & \B{BoolOp} \\
        \B{args} & : & seq \B{Pred} \\
        \hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{Struct} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Data structure : \B{Struct}} \\
        \hline
	\B{id} & : &  \B{Name}\\
        \B{elements} & : & seq \B{Name} \\
	\B{type} & : & seq \B{Type}\\
	\hline
%  	\multicolumn{3}{|c|}{Data record : \B{Record}} \\
%        \hline
%	\B{id} & : &  \B{Name}\\
%        \B{elements} & : & seq \B{Name} \\
%	\B{type} & : & \B{Struct}\\
%	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Expr}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Expr} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).
      Sequence and optional attributes are denoted ``seq'' and ``opt'',
      respectively}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

The rule lvalue is:
\begin{align*}
\lefteqn{\trad{n}{lvalue} \isdef} \\
& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}}"} \IN \\
& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
& \quad \IF \mbox{\B{n} is an array} \THEN  \trad{n}{ArrayItem} \\
& \quad \ELSE \LET v = \variable \IN \\
& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
\end{align*}

@@

A list of type and variable is generated by the following rule.
%TODO: insert a rule to do the shift address with the first element of interval
\begin{align*}
&\trad{n}{LRExp} \isdef \LET p \sep v \sep t  = \trad{n}{ShiftExp}  \IN  \llvm{",\PH{t} \PH{v} "} \sep p\\
&\trad{n l}{LRExp} \isdef \LET r_1 \sep p_1 = \trad{n}{LRExp} \AND 
 \LET r_2 \sep p_2 = \trad{l}{LRExp} \IN\\
& \quad  \llvm{"\PH{r_1} \PH{r_2}"} \sep \llvm{"\PH{p_1} \PH{p_2}"}\\
\end{align*}

The element access of array can need a shift in indexes. When the domain of 
array is an interval and it is not started from zero, it must adjust the 
indexes shifting the access, because the addressing  of array in intermediate 
representation LLVM starts at 0. To access an element from array the formula is 
$address(e(i1,..,in))  = i1-i1\_start, ... , in-in\_start$. The following rule 
\footnote{The function $TSymbol$ returns the "start", "end" and type of the one 
index.} makes the shift for each index of array. \VGM{TODO: The function TSymbol
returns the "start", "end" and type of the one index.}

\begin{align*}
&\trad{n}{ShiftExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp} \AND \\
&\quad iStart \sep iEnd \sep iType = \trad{n}{TSymbol}  \AND  \ell = \lbl \IN \\
&\quad \IF ( iStart == 0 ) \THEN  \llvm{",\PH{t} \PH{v} "} \sep p \\
&\quad \ELSE \LET  p \sep v \sep v' \sep t'  = \trad{p,t,v,itype,istart}{Cast} \IN \\
&\quad \llvm{",\PH{t} \PH{l} "} \sep  \llvm{"\PH{p} \PH{\ell} = sub \PH{t'} \PH{v}, \PH{v'}\nl"}  \\
\end{align*}

The \trad{}{ShiftExp} rule does a subtraction and this operation must have operators 
with the same data types so the following cast rule is used:

\begin{align*}
&\trad{p,t,v,t',v'}{Cast} \isdef \LET  \ell = \lbl \IN \\
&\quad \IF ( t == t') \THEN p \sep v \sep v' \sep t  \\
&\quad \ELSE \IF (t<t') \THEN \\
&\quad \quad \llvm{"\PH{p} \PH{l}= bitcast \PH{t} \PH{v} , \PH{t'}\nl"}  \sep l \sep v' \sep t'\\ 
&\quad \ELSE \\ 
&\quad \quad \llvm{"\PH{p} \PH{l}= bitcast \PH{t'} \PH{v'} , \PH{t}\nl"} \sep v \sep l \sep t \\ 
\end{align*}
